Staged Abstract Interpreters

### 2019-09-06 paper/10_intro.tex
typo
revision

However, besides the intrinsic complexity of static analysis, there are
additional abstraction penalties from these high-level implementation
approaches. First, similar to concrete interpreters, an abstract interpreter
analyzes a program by traversing its AST. This traversal incurs
interpretive overhead, such as pattern matching on the AST nodes and recursive
calls on subexpressions. If it is a single-time traversal, such overhead can be
negligible. But usually, abstract interpreters iteratively analyze and traverse
the AST multiple times to reach fixed-points, therefore the accumulated
overhead is even magnified. 
%Repeatedly analyzing library programs also poses significant overhead.
%Such overhead can be negligible if the abstract
%interpreter only runs on the program a single time, but can accumulate
%significantly if it runs repeatedly (for example, on libraries).  
Second, it is common that abstract interpreters written in pure languages
extensively use functional encodings of side-effects to implement the semantics
of abstract interpretation.  For example, the list monad can be used to model
nondeterminism from collecting interpreters.
%For example, a collecting interpreter that returns all possible runtime values is
%nondeterministic, which can be modeled by a list monad.
Although such pure approaches have their own merits, compared with imperative,
stateful implementations, they are significantly slower.  In order to
reconcile these approaches, one must use a programming technique which allows
for high-level abstractions during implementation, but bypasses the
incurred overhead at runtime.

Roughly at the same time when abstract interpretation was proposed in the
1970s, \citet{futamura1971partial} observed a close
connection between interpreters and compilers through a hierarchy of
specializations, since then known as Futamura Projections. The first Futamura
projection describes that specializing an interpreter to the input program
removes interpretative overhead and yields an equivalent executable
that runs faster. The Futamura projections have been successfully applied to
other similar performance problems, examples including
query compilers \cite{DBLP:conf/sigmod/TahboubER18}, PyPy \footnote{PyPy:
\url{http://pypy.org/}}, Truffle \cite{Marr:2015:TVP:2814270.2814275}, etc.
The challenge here: can we apply the same idea to a general abstract interpreter
to make it run faster? The solution is not obvious, as we need to specialize
a nonstandard semantics that extensively uses nondeterminism, operates on
abstract domains, and computes fixed-points iteratively over a cache.

>>>>

However, besides the intrinsic complexity of static analysis, there are
additional abstraction penalties from these high-level implementation
approaches. First, similar to concrete interpreters, an abstract interpreter
analyzes a program by traversing its AST. This traversal incurs
interpretive overhead, such as pattern matching on the AST nodes and recursive
calls on subexpressions. If it is a single-time traversal, such overhead can be
negligible. But usually, abstract interpreters iteratively analyze and traverse
the AST multiple times to reach fixed-points, therefore the accumulated
overhead is even magnified. 
%Repeatedly analyzing library programs also poses significant overhead.
%Such overhead can be negligible if the abstract
%interpreter only runs on the program a single time, but can accumulate
%significantly if it runs repeatedly (for example, on libraries).  
Second, it is common that abstract interpreters written in pure languages
extensively use functional encodings of side-effects to implement the semantics
of abstract interpretation.  For example, the list monad can be used to model
nondeterminism from collecting interpreters.
Although such pure approaches have their own merits, compared with imperative,
stateful implementations, they are significantly slower.  In order to
reconcile these approaches, one must use a programming technique which allows
for high-level abstractions during implementation, but bypasses the
incurred overhead at runtime.

Roughly at the same time when abstract interpretation was proposed in the
1970s, \citet{futamura1971partial} observed a close
connection between interpreters and compilers through a hierarchy of
specializations, since then known as Futamura Projections. The first Futamura
projection describes that specializing an interpreter to the input program
removes interpretative overhead and yields an equivalent executable
that runs faster. 
In recent years, the idea of Futamura projection has regained more attentions
and successfully applied to many real-world problems, examples including
query compilers \cite{DBLP:conf/sigmod/TahboubER18}, PyPy \cite{Bolz:2009:TMP:1565824.1565827},
Truffle \cite{Marr:2015:TVP:2814270.2814275}, etc.
The challenge here: can we apply the same idea to a general abstract interpreter
to make it run faster? The solution is not obvious, as we need to specialize
a nonstandard semantics that extensively uses nondeterminism, operates on
abstract domains, and computes fixed-points iteratively over a cache.
